---
title: "Basic Inferential Data Analysis"
author: "Daniel Cevallos"
date: "21/06/2023"
output:
  html_document: default
  pdf_document: default
---

# Overview
We're going to analyze the ToothGrowth data in the R datasets package.

1. Load the ToothGrowth data and perform some basic exploratory data analyses
2. Provide a basic summary of the data.
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there’s other approaches worth considering)
4. State your conclusions and the assumptions needed for your conclusions.


# 1. Load the ToothGrowth data and perform some basic exploratory data analyses
```{r}
data("ToothGrowth")
dt <- ToothGrowth
```

# 2. Provide a basic summary of the data.


```{r}
str(dt)
```
The data has 60 rows and 3 columns

```{r}
unique(dt$dose)
```

Since there are three unique values for dose, I will convert it to a factor type.

```{r}
dt$dose <- factor(dt$dose)
str(dt)
```

Using summary function to look at statistics of variables
```{r}
summary(dt)
```

Plotting box plot based on supplement type and dose
```{r}
library(ggplot2)
ggplot(dt, aes(x = dose, y = len, fill = supp)) +
    geom_boxplot() +
    ggtitle("Length of tooth growth based on supplement type and dose") +
    theme_classic()
```
 
From the box plot, we can observe that the dosage for 0.5 and 1.0 has bigger differences in length as compared to the dosage for 2.0 mg/day. We can see a trend as the dose increases, the tooth length increases as well. Moreover, from the plot alone, we can tell supplement OJ is more effective for doses 0.5 and 1.0, where as for dose 2, there's not much difference between them.

\newpage

We can look at the mean for the length of tooth growth for each dose and supp.

load libraries 
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
```

```{r}
dt %>%
    group_by(supp, dose) %>%
    summarize(mean = mean(len), .groups='drop') %>%
    spread(supp, mean) %>%
    mutate(diff = abs(VC - OJ))
```

Observing the mean, we can see 2mg/day doses has very small differences as compared to dose of 0.5 and 2. This means it's harder to compare the effectiveness between OJ and VC for dose 2. 

To formally test the effectiveness between the two supplement types, we will use t.test to find the p-values and confidence intervals to perform hypothesis testing.


# 3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.

* Our null hypothesis would is there is no difference between using OJ and VC
* Our alternative hypothesis is there is a difference between sign OJ and VC
* alpha rate is set at 0.05 as standard  

Even though dose 2 has the smallest differences, I will still be performing hypothesis testing for all three dose types. 

First we filter the data based on dose. This makes it easier and cleaner for t.test.
```{r}
dose_half <- filter(dt, dose == 0.5)
dose_one <- filter(dt, dose == 1)
dose_two <- filter(dt, dose == 2)
```

### t-test for 0.5 mg/day dose
```{r}
t.test(len ~ supp, dose_half)
```

###  t-test for 1 mg/day dose

```{r}
t.test(len ~ supp, dose_one)
```

###  t-test for 2 mg/day dose

```{r}
t.test(len ~ supp, dose_two)
```

\pagebreak

Forming a table to summarize the t.tests results 

```{r}
dose <- c(0.5, 1.0, 2.0)
p_value <- c(0.0064, 0.0010, 0.9639)
conf.int <- c("1.72, 8.78", "2.80, 9.06", "-3.80, 3.64")
decision <- c("Reject null", "Reject null", "Do not reject null")
data.frame(dose, conf.int, p_value, decision)
```

As expected, the p-values for dose 0.5 and 1.0 will be very small because of the big differences in mean between them. 

Thus, for dose 0.5 and 1.0, since p-values are smaller than 0.5, we reject the null hypotheses that the supplement types don't have a difference on tooth growth. But for dose 2.0 mg/day, we can reject the null as the p-value is greater than 0.5.

# 4. State your conclusions and the assumptions needed for your conclusions.

The central assumption for the results is that the sample is representative of the population, and the variables are IID random variables. 

For the t.test, two assumptions are made,

1. The data isn't paired, meaning they're independent
2. The variance are different.

With 95% confidence we reject the null hypothesis, stating that there is no difference in the tooth growth by the delivery method for .5 and 1 milligrams/day. We observe p-values less than the treshold of .05 and the confidence levels don’t include 0. So, for dosage of .5 milligrams/day and 1 milligrams/day does matter the delivery method. With 95% confidence we fail to reject the null hypothesis, stating that there is no difference in the tooth growth by the delivery method for 2 milligrams/day. We observe p-values more than the treshold of .05 and the confidence levels include 0. So, for dosage of 2 milligrams/day the delivery method doesn’t matter. 